\chapter{Literature Review}\thispagestyle{empty}
\graphicspath{{Chapter2/figures/}}
{\setstretch{1.5}

The application of artificial intelligence to lung cancer detection and diagnosis has grown
substantially in recent years, driven by the clinical urgency of early detection and the
availability of large annotated CT imaging datasets. This chapter reviews fifteen relevant
works spanning CT-based deep learning, explainable AI, multimodal fusion, and pulmonary
nodule detection, organised thematically to contextualise the design decisions underlying
LungInsight.

%-----------------------------------------------------------------------
\section{CNN-Based Detection from CT Imaging}

The most extensively explored paradigm in lung cancer AI involves training convolutional
neural networks directly on CT scan images. Hammad et al. \cite{hammad2025} proposed a
custom CNN augmented with Grad-CAM explainability for lung cancer detection from CT
images, achieving a classification accuracy of approximately 95\% with visual
localisation of tumour regions. While the model demonstrated strong discriminative
performance, it was trained on a single dataset and showed limited generalisability
across different scanner types and patient populations, with no integration into clinical
workflows or longitudinal outcome tracking.

Murthy et al. \cite{murthy2025} pursued a real-time diagnostic objective through an
optimised CNN architecture, achieving fast inference suitable for clinical deployment.
However, the study did not report standardised performance metrics or compare results
against radiologist baselines, making reproducibility and clinical benchmarking
difficult. Similarly, Prathima et al. \cite{prathima2025} conducted a comprehensive
comparative study of CNN and ensemble deep learning models on CT scans, reporting
accuracies in the range of 93--97\% with balanced sensitivity and specificity. Despite
the robustness of the comparative design, the study's high computational cost and
absence of early-stage focus limit its practical clinical applicability.

Achyutha Sree et al. \cite{achyuthasree2025} proposed a multi-objective deep learning
pipeline on the LIDC-IDRI low-dose CT dataset, achieving accuracies above 90\% with
improved PSNR and SSIM metrics and reduced false positives. The work highlighted real-time
feasibility challenges and limited dataset diversity as key barriers to deployment.
Gondkar et al. \cite{gondkar2024} introduced a hybrid CNN-GRU architecture combining
spatial and sequential learning on CT scans, reporting accuracies of 92--95\%. However,
the use of GRU layers for inherently static image data was identified as architecturally
redundant, and the small dataset size constrained the generalisability of results.

%-----------------------------------------------------------------------
\section{Explainable AI in Lung Cancer Diagnosis}

Recognising that black-box predictions are insufficient for clinical trust, several
works have incorporated explainability into their diagnostic pipelines.
Alqahtani et al. \cite{alqahtani2025} applied the EfficientNet architecture with
Grad-CAM visualisation for automated lung cancer stage classification, achieving
accurate stage prediction accompanied by interpretable heatmaps. The model was,
however, restricted to three output classes (normal, malignant, benign), with
insufficient handling of early-stage or rare subtypes and no evaluation of multimodal
data fusion.

Bouamrane et al. \cite{bouamrane2024} addressed model robustness by combining curriculum
learning, deep CNNs, and explainable AI across multiple CT datasets, demonstrating
improved cross-dataset performance and generalisation. The approach remained limited to
CT imaging alone, with scalability challenged by heterogeneous preprocessing requirements
and no assessment of false-positive rates across diverse demographic groups.

Zheng et al. \cite{zheng2021} presented an interpretable computer-aided diagnosis system
integrating radiology images with clinical data for malignancy evaluation, with a clear
emphasis on clinical usability. While interpretability was central to the design, the
work predates current foundation model developments, relies on a single dataset, and does
not account for comorbidities or three-dimensional CT dynamics.

Chaudhari and Gudadhe \cite{chaudhari2025} conducted a survey of AI methods applied to
lung cancer diagnosis and treatment, providing a broad overview of machine learning and
deep learning trends in the domain. The work was, however, non-experimental in nature,
offering no quantitative benchmarking and no guidance on deployment in low-resource
settings — a gap the authors themselves identified as needing urgent attention.

%-----------------------------------------------------------------------
\section{Multimodal Fusion Approaches}

A growing body of work has sought to move beyond single-modality imaging by incorporating
structured clinical data alongside CT scans. Oncu and Ciftci \cite{oncu2025} proposed a
multimodal AI framework integrating CNN-extracted image features with ANN-processed
clinical data, demonstrating improved diagnostic accuracy over imaging-only baselines.
The study did not specify the datasets used or validation metrics applied, and gaps in
clinical data standardisation and handling of missing modalities remain unresolved.

Yamada et al. \cite{yamada2024} explored multimodal deep learning by combining low-dose
CT scans with structured clinical information for the diagnosis of mediastinal tumours,
demonstrating the radiation-reducing benefit of low-dose imaging. The work was limited
to mediastinal rather than primary lung parenchymal pathology, and the small preliminary
cohort introduced bias risks with no long-term outcome data.

Niu et al. \cite{niu2025} introduced a medical multimodal multitask foundation model for
lung cancer screening that handles both imaging and clinical data at scale. While the
architecture demonstrated strong performance across multiple screening tasks, the model
remained focused on screening rather than staging or treatment recommendation, and its
dependency on high-quality annotations constrained applicability in under-resourced
settings. Interpretability for non-expert clinical users was also identified as a
significant limitation.

%-----------------------------------------------------------------------
\section{Pulmonary Nodule Detection}

Nodule-level detection forms a foundational component of any lung cancer diagnostic
pipeline. Salhi et al. \cite{salhi2026} applied AI-based detection and classification
models to the widely used LIDC-IDRI CT dataset, achieving improved pulmonary nodule
detection accuracy. The study relied on legacy LIDC-IDRI annotations without
incorporating more recent expert consensus annotations, and did not address nodule
malignancy risk scoring, which is clinically essential for treatment decisions.

Koyyada et al. \cite{koyyada2023} proposed a CNN combined with object detection for
nodule identification in chest X-rays rather than CT scans, reporting accuracy in the
range of 85--90\% with IoU-based localisation metrics. While the approach offers a
low-cost, accessible alternative to CT-based methods, X-ray modality inherently limits
sensitivity for small nodules and introduces higher false-negative rates compared to CT
imaging.

%-----------------------------------------------------------------------
\section{Survey of NSCLC Detection Methods}

Pawar et al. \cite{pawar2025} reviewed multiple datasets and AI methods applied to
non-small cell lung cancer detection, reporting accuracy figures up to 97\% across
surveyed studies. As a review paper without experimental validation, the work provided
broad directional insights but lacked deployment guidance and real-time applicability
discussion — consistent with the broader pattern of a field rich in benchmarked models
but sparse in clinically deployed systems.

%-----------------------------------------------------------------------
\section{Research Gaps and Positioning of LungInsight}

Across the reviewed literature, several consistent gaps emerge. First, the majority of
high-performing models rely exclusively on CT imaging, discarding the complementary
diagnostic signal available in structured clinical records such as smoking history, age,
emphysema score, and lung function. Second, explainability — where present — is typically
limited to imaging heatmaps via Grad-CAM, with few systems offering feature-level
attribution for clinical variables. Third, most works are evaluated on a single dataset,
limiting generalisability across populations and scanner types. Finally, end-to-end
clinical deployment with an accessible interface for non-specialist practitioners remains
largely unaddressed.

LungInsight addresses these gaps by developing a multimodal late-fusion architecture that
jointly encodes CT imaging through a DenseNet-based convolutional branch and clinical
tabular data through an MLP-based encoder. Explainability is applied at both modalities:
Grad-CAM localises suspicious regions in CT slices, while SHAP values quantify the
contribution of individual clinical features to each patient-level prediction. The system
is evaluated using a comprehensive suite of metrics — including ROC-AUC, F1-score,
PR-AUC, Brier score, and calibration curves — and is delivered through a Streamlit
dashboard that enables real-time clinical use without specialist AI expertise.

%-----------------------------------------------------------------------
\begin{table}[H]
\centering
\captionsetup{justification=centering}
\caption{Summary of Reviewed Literature}
\resizebox{\textwidth}{!}{
\begin{tabular}{clllll}
\toprule
\textbf{SI No.} & \textbf{Author(s)} & \textbf{Year} & \textbf{Method} & \textbf{Key Result} & \textbf{Key Gap} \\
\midrule
1 & Hammad et al. & 2025 & Custom CNN + Grad-CAM & $\sim$95\% accuracy & Single dataset; no clinical integration \\
2 & Alqahtani et al. & 2025 & EfficientNet + Grad-CAM & Accurate stage classification & 3-class only; no multimodal fusion \\
3 & Chaudhari \& Gudadhe & 2025 & Survey of ML/DL methods & Overview of AI applications & No quantitative benchmarking \\
4 & Murthy et al. & 2025 & Optimised CNN & Fast inference & No standard metrics or radiologist baseline \\
5 & Bouamrane et al. & 2024 & CNN + Curriculum Learning + XAI & Improved cross-dataset robustness & CT only; no multimodal inputs \\
6 & Niu et al. & 2025 & Multimodal foundation model & Strong screening performance & Screening only; limited interpretability \\
7 & Zheng et al. & 2021 & Interpretable CAD & Improved malignancy prediction & Outdated; single dataset \\
8 & Oncu \& Ciftci & 2025 & CNN + ANN (multimodal) & Improved diagnostic accuracy & Unspecified datasets; missing modality gaps \\
9 & Salhi et al. & 2026 & AI on LIDC-IDRI & High nodule detection accuracy & Legacy annotations; no risk scoring \\
10 & Yamada et al. & 2024 & Multimodal DL (low-dose CT) & Promising mediastinal results & Mediastinal scope only; small cohort \\
11 & Achyutha Sree et al. & 2025 & Multi-objective DL on LDCT & $>$90\% accuracy & High complexity; limited dataset diversity \\
12 & Gondkar et al. & 2024 & CNN + GRU & 92--95\% accuracy & GRU redundant for static images \\
13 & Prathima et al. & 2025 & CNN / Ensemble & 93--97\% accuracy & High cost; no early-stage focus \\
14 & Koyyada et al. & 2023 & CNN + Object Detection (X-ray) & 85--90\% accuracy & Lower than CT; high false negatives \\
15 & Pawar et al. & 2025 & Survey (NSCLC) & Up to 97\% across studies & No experimental validation \\
\bottomrule
\end{tabular}}
\label{LitSurveyTable}
\end{table}

}