\chapter{Supplementary Dataset and Preprocessing Details}
\thispagestyle{empty}

This appendix supplements Chapter~4 by documenting the complete dataset
specifications, patient-level matching methodology, and preprocessing steps
used to construct the multimodal corpus that drives the LungInsight pipeline.
The intent is to provide enough detail for independent reproduction of the
data layer described in the main thesis.

%-----------------------------------------------------------------------
\section{Data Sources}
\label{sec:data-sources}

Two publicly available, IRB-approved repositories form the basis of the
LungInsight corpus. Imaging data are drawn from the
\textbf{LIDC-IDRI} collection~\cite{armato2011lung}, distributed through the
Cancer Imaging Archive (TCIA)~\cite{clark2013cancer}. Structured clinical
covariates are drawn from the \textbf{National Lung Screening Trial
(NLST)}~\cite{national2011reduced} clinical dataset, accessed under the
applicable data-use agreement.

\subsection{LIDC-IDRI}
LIDC-IDRI contains 1{,}018 thoracic CT cases with nodule-level annotations
contributed by up to four expert radiologists per case. Each nodule is
annotated with a bounding box, a pixel-level segmentation contour, and a
malignancy likelihood rating on a five-point scale. We restrict the cohort to
cases with at least three concurring annotators and use the median malignancy
rating as the consensus label.

\subsection{NLST Clinical Records}
NLST provides demographic, behavioural, physiological, and outcome variables
for 53{,}454 participants enrolled in the original screening trial. For
LungInsight the selected variables are:

\begin{itemize}
    \item age (years),
    \item sex,
    \item pack-year smoking history,
    \item smoking status (current, former, never),
    \item emphysema score (radiologist-graded),
    \item spirometric lung function indices (FEV$_1$, FVC, FEV$_1$/FVC),
    \item family history of lung cancer,
    \item ground-truth diagnostic outcome (benign, malignant, malignancy stage).
\end{itemize}

%-----------------------------------------------------------------------
\section{Patient-Level Matching}
\label{sec:patient-matching}

Because LIDC-IDRI and NLST are independently collected cohorts, patient-level
fusion is performed via a probabilistic matching procedure rather than a
direct record linkage. Each candidate (CT case, NLST record) pair is scored
on the following criteria:

\begin{itemize}
    \item age difference within $\pm 1$ year,
    \item identical reported sex,
    \item identical smoking-status category (current, former, never),
    \item agreement on reported emphysema grade or lung-function impairment
          band,
    \item nodule size similarity, when nodule descriptors are available in
          both sources.
\end{itemize}

A confidence score $c \in [0, 1]$ is computed for every candidate pair as a
weighted sum of the per-criterion agreements. Pairs with $c < 0.75$ are
discarded to reduce false matches and prevent label noise. After filtering,
the paired cohort contains $847$ matched cases out of the original $1{,}018$
LIDC-IDRI scans.

%-----------------------------------------------------------------------
\section{Imaging Preprocessing}
\label{sec:imaging-preproc}

The CT preprocessing pipeline converts raw DICOM series into model-ready
patches via the following deterministic steps:

\begin{enumerate}
    \item Convert each DICOM volume to NIfTI format using
          \texttt{SimpleITK}~\cite{muschelli2014simpleitk}.
    \item Apply lung windowing with centre $-600$\,HU and width $1500$\,HU to
          isolate pulmonary parenchyma.
    \item Clip voxel intensities to the range $[-1000, 400]$\,HU.
    \item Min--max normalise clipped values to $[0, 1]$.
    \item Crop $64 \times 64$ axial patches centred on annotated nodule
          centroids with a $5$\,mm contextual margin.
\end{enumerate}

During training, patches are augmented on-the-fly with random horizontal and
vertical flips, rotations of up to $\pm 15^{\circ}$, and Gaussian intensity
jitter with $\sigma = 0.02$. Augmentation is disabled during validation and
test inference.

\subsection{Patch Cropping Algorithm}
\label{sec:patch-crop}

Each patch is extracted from the axial slice containing the nodule centroid
together with its immediate spatial neighbourhood. The cropping routine pads
boundary patches symmetrically to ensure a fixed $64 \times 64$ output
regardless of nodule location:

\begin{verbatim}
for each nodule centroid (x0, y0) on slice S:
    half  = patch_size // 2
    x_min = max(0, x0 - half)
    x_max = min(width,  x0 + half)
    y_min = max(0, y0 - half)
    y_max = min(height, y0 + half)
    patch = S[y_min:y_max, x_min:x_max]
    if patch.shape != (64, 64):
        pad patch symmetrically with the slice mean
\end{verbatim}

This guarantees that every patch presented to the imaging branch has the same
spatial dimensions, allowing batched GPU processing without per-sample
reshaping.

%-----------------------------------------------------------------------
\section{Clinical Preprocessing}
\label{sec:clinical-preproc}

Clinical covariates are preprocessed as follows:

\begin{itemize}
    \item \textbf{Continuous variables} (age, pack-years, FEV$_1$, FVC,
          FEV$_1$/FVC) are standardised to zero mean and unit variance using
          statistics computed on the training split only, to prevent
          information leakage from validation or test sets.
    \item \textbf{Categorical variables} (sex, smoking status, family history,
          emphysema grade) are one-hot encoded.
    \item \textbf{Missing values}, present in approximately 4\% of clinical
          records, are imputed using the training-set median for continuous
          features and the training-set mode for categorical ones.
    \item The final clinical feature vector has a fixed length of $18$ and is
          assembled in a canonical column order so that SHAP attributions in
          Appendix~\ref{ch:impl-details} can be interpreted consistently.
\end{itemize}

%-----------------------------------------------------------------------
\section{Data Splits}
\label{sec:data-splits}

The matched paired cohort is partitioned into training, validation, and test
splits with proportions of approximately $70\%$, $15\%$, and $15\%$
respectively. Stratified sampling on the binary malignancy label ensures that
class proportions are preserved across all three splits. A strict
no-patient-overlap constraint is enforced so that no patient appears in more
than one partition, eliminating leakage of patient-specific imaging or
clinical signatures across splits.

Table~\ref{tab:appendix1-splits} summarises the resulting split sizes.

\begin{table}[H]
\centering
\caption{LungInsight data splits after patient-level matching and stratified
sampling.}
\label{tab:appendix1-splits}
\begin{tabular}{lccc}
\toprule
\textbf{Split} & \textbf{Cases} & \textbf{Malignant (\%)} & \textbf{Benign (\%)} \\
\midrule
Training   & 593 & 32 & 68 \\
Validation & 127 & 31 & 69 \\
Test       & 127 & 33 & 67 \\
\midrule
\textbf{Total} & \textbf{847} & \textbf{32} & \textbf{68} \\
\bottomrule
\end{tabular}
\end{table}

All random seeds used for cohort partitioning, weighted sampling, and
augmentation are logged alongside the model checkpoints to enable exact
replication of every reported experimental result.