\begin{thebibliography}{99}
\bibitem{armato2011lung} S. G. Armato III, G. McLennan, L. Bidaut, M. McNitt-Gray, C. R. Meyer, A. P. Reeves, B. R. Yankelevitz, E. A. Henschke, D. R. Aberle, C. I. Moskowitz, and C. S. Clarke, ``The Lung Image Database Consortium (LIDC) and Image Database Resource Initiative (IDRI): A completed reference database of lung nodules on CT scans,'' \emph{Medical Physics}, vol. 38, no. 2, pp. 915--931, Feb. 2011.

\bibitem{national2011reduced} N. B. National Lung Screening Trial Research Team, ``Reduced lung-cancer mortality with low-dose computed tomographic screening,'' \emph{New England Journal of Medicine}, vol. 365, no. 5, pp. 395--409, 2011.

\bibitem{clark2013cancer} K. Clark, B. Vendt, K. Smith, J. Freymann, J. Kirby, P. Koppel, S. Moore, S. Phillips, D. Maffitt, M. Pringle, M. Tarbox, and F. Prior, ``The Cancer Imaging Archive (TCIA): Maintaining and operating a public information repository,'' \emph{Journal of Digital Imaging}, vol. 26, no. 6, pp. 1045--1057, 2013.

\bibitem{huang2017densely} G. Huang, Z. Liu, L. Van Der Maaten, and K. Q. Weinberger, ``Densely connected convolutional networks,'' in \emph{Proceedings of the IEEE Conference on Computer Vision and Pattern Recognition (CVPR)}, 2017, pp. 4700--4708.

\bibitem{selvaraju2017grad} R. R. Selvaraju, M. Cogswell, A. Das, R. Vedantam, D. Parikh, and D. Batra, ``Grad-CAM: Visual explanations from deep networks via gradient-based localization,'' in \emph{Proceedings of the IEEE International Conference on Computer Vision (ICCV)}, 2017, pp. 618--626.

\bibitem{lundberg2017unified} S. M. Lundberg and S.-I. Lee, ``A unified approach to interpreting model predictions,'' in \emph{Advances in Neural Information Processing Systems (NeurIPS)}, 2017, pp. 4765--4774.

\bibitem{lin2017focal} T.-Y. Lin, P. Goyal, R. Girshick, K. He, and P. Dollár, ``Focal loss for dense object detection,'' in \emph{Proceedings of the IEEE International Conference on Computer Vision (ICCV)}, 2017, pp. 2980--2988.

\bibitem{loshchilov2019decoupled} I. Loshchilov and F. Hutter, ``Decoupled weight decay regularization,'' in \emph{International Conference on Learning Representations (ICLR)}, 2019.

\bibitem{paszke2019pytorch} A. Paszke, S. Gross, F. Massa, A. Lerer, J. Bradbury, G. Chanan, T. Killeen, Z. Lin, N. Gimelshein, L. Antiga, A. Desmaison, A. Kopf, E. Yang, Z. DeVito, M. Raison, A. Tejani, S. Chilamkurthy, B. Steiner, L. Fang, J. Bai, and S. Chintala, ``PyTorch: An imperative style, high-performance deep learning library,'' in \emph{Advances in Neural Information Processing Systems (NeurIPS)}, 2019, pp. 8024--8035.

\bibitem{pedregosa2011scikit} F. Pedregosa, G. Varoquaux, A. Gramfort, V. Michel, B. Thirion, O. Grisel, M. Blondel, P. Prettenhofer, R. Weiss, V. Dubourg, J. Vanderplas, A. Passos, D. Cournapeau, M. Brucher, M. Perrot, and E. Duchesnay, ``Scikit-learn: Machine learning in Python,'' \emph{Journal of Machine Learning Research}, vol. 12, pp. 2825--2830, 2011.

\bibitem{muschelli2014simpleitk} J. Muschelli, N. J. Johnson, K. R. M. V. J., and D. C. Marcus, ``SimpleITK image-analysis notebooks: A collaborative environment for education and reproducible research,'' \emph{Journal of Digital Imaging}, vol. 27, no. 3, pp. 411--420, 2014.

\bibitem{streamlit2021} Streamlit Inc., ``Streamlit,'' 2021. [Online]. Available: https://streamlit.io

\bibitem{docker2020} Docker Inc., ``Docker,'' 2020. [Online]. Available: https://www.docker.com
\end{thebibliography}
