\chapter{Implementation and Experimental Details}
\label{ch:impl-details}
\thispagestyle{empty}

This appendix supplements Chapter~4 by documenting the implementation
choices, training configuration, explainability setup, and software
environment used to realise the LungInsight system. The level of detail is
chosen to be sufficient for an independent team to reproduce the results
reported in the main thesis.

%-----------------------------------------------------------------------
\section{Model Implementation}
\label{sec:appB-model}

The LungInsight model is implemented as a single PyTorch~\cite{paszke2019pytorch}
\texttt{nn.Module} with two parallel branches and a late-fusion head.

\begin{itemize}
    \item A \textbf{DenseNet-121}~\cite{huang2017densely} imaging branch
          pretrained on ImageNet, adapted for single-channel grayscale CT
          input by replacing the first convolutional layer with a
          channel-averaged copy of the original RGB filters. The final
          classification head is replaced by a $256$-dimensional fully
          connected projection layer with ReLU activation and batch
          normalisation, producing an imaging embedding
          $\mathbf{z}_{\text{img}} \in \mathbb{R}^{256}$.
    \item A \textbf{three-layer clinical multilayer perceptron} that maps the
          $18$-dimensional clinical vector to a $64$-dimensional embedding
          $\mathbf{z}_{\text{clin}} \in \mathbb{R}^{64}$. Each hidden layer
          uses ReLU activation, batch normalisation, and dropout with
          probability $0.3$.
\end{itemize}

The two embeddings are concatenated into a $320$-dimensional fused
representation $\mathbf{z} = [\mathbf{z}_{\text{img}} \,\|\, \mathbf{z}_{\text{clin}}]$.
Two output heads operate on this representation:

\begin{itemize}
    \item a \textbf{binary malignancy head} with sigmoid activation producing
          $p(\text{malignant} \mid \mathbf{z}) \in [0, 1]$;
    \item a \textbf{four-class stage head} with softmax activation producing
          $p(\text{stage} \mid \mathbf{z}) \in \Delta^{3}$.
\end{itemize}

%-----------------------------------------------------------------------
\section{Training Configuration}
\label{sec:appB-training}

The model is trained end-to-end with the configuration summarised in
Table~\ref{tab:appendix2-training}. Class imbalance is addressed both at the
sampling level (weighted random sampler) and at the loss level (focal
loss~\cite{lin2017focal} with inverse class-frequency weighting). Optimisation
uses AdamW~\cite{loshchilov2019decoupled} with decoupled weight decay.

\begin{table}[H]
\centering
\caption{LungInsight training configuration.}
\label{tab:appendix2-training}
\begin{tabular}{l p{0.62\textwidth}}
\toprule
\textbf{Hyperparameter} & \textbf{Value} \\
\midrule
Optimiser                       & AdamW with weight decay $10^{-4}$~\cite{loshchilov2019decoupled} \\
Backbone learning rate          & $1 \times 10^{-4}$ \\
Projection / head learning rate & $5 \times 10^{-4}$ \\
Batch size                      & $16$ paired (CT, clinical) examples \\
Epochs                          & $60$ (early stopping at $\sim$45) \\
LR scheduler                    & Cosine annealing with warm restarts \\
Warm-up strategy                & Backbone frozen for first $3$ epochs \\
Loss (malignancy)               & Focal loss, $\gamma = 2$, $\alpha$ from class frequency \\
Loss (stage)                    & Weighted cross-entropy \\
Mixed precision                 & Enabled via \texttt{torch.cuda.amp} \\
Gradient clipping               & $\ell_{2}$ norm clipped at $1.0$ \\
Random seed                     & $42$ (cohort split, sampler, augmentation) \\
\bottomrule
\end{tabular}
\end{table}

Early stopping monitors validation ROC-AUC with a patience of $10$ epochs.
The checkpoint with the best validation ROC-AUC is retained for evaluation on
the held-out test split.

%-----------------------------------------------------------------------
\section{Explainability and Inference}
\label{sec:appB-xai}

\textbf{Grad-CAM.} Grad-CAM~\cite{selvaraju2017grad} is applied to the final
convolutional block of the DenseNet imaging branch
(\texttt{model.imaging\_branch.features.denseblock4}). For a given CT patch,
the gradient of the predicted malignancy score with respect to the target
feature maps is global-average-pooled to produce channel weights; the
weighted activation map is upsampled bilinearly to the input patch resolution
and normalised to $[0, 1]$. During inference, all batch-normalisation layers
are placed in evaluation mode and the Grad-CAM computation is performed
within a \texttt{torch.no\_grad()} context to avoid retaining unnecessary
computation graphs.

\begin{verbatim}
from pytorch_grad_cam import GradCAM
from pytorch_grad_cam.utils.model_targets import ClassifierOutputTarget

target_layer = model.imaging_branch.features.denseblock4
cam          = GradCAM(model=model, target_layers=[target_layer])
heatmap      = cam(input_tensor=ct_patch,
                   targets=[ClassifierOutputTarget(1)])
\end{verbatim}

\textbf{SHAP.} SHAP~\cite{lundberg2017unified} values are computed using a
\texttt{DeepExplainer} instantiated with a background set of $100$
representative training clinical feature vectors. Separate attribution sets
are produced for: (i) the raw clinical inputs, and (ii) the fused
representation $\mathbf{z}$. The clinical-input attributions are rendered as
SHAP waterfall plots in the dashboard; the cohort-level attributions are
rendered as beeswarm summary plots.

\textbf{Inference pipeline.} A dedicated \texttt{InferencePipeline} class
wraps the model and exposes a single
\texttt{predict(dicom\_path, clinical\_dict)} interface that performs
DICOM-to-tensor conversion, clinical vector construction, model forward pass,
Grad-CAM, and SHAP computation in sequence. Model weights are exported to
TorchScript for faster dashboard inference, and SHAP background samples are
precomputed on application start-up.

%-----------------------------------------------------------------------
\section{Dashboard and Deployment}
\label{sec:appB-dashboard}

The clinical decision-support interface is implemented with
Streamlit~\cite{streamlit2021}. The dashboard accepts either a DICOM CT
series upload or a CSV file containing clinical covariates, and displays:

\begin{itemize}
    \item the predicted malignancy probability rendered as a gauge chart;
    \item the predicted disease stage with per-class confidence scores;
    \item a Grad-CAM heatmap overlaid on the most informative axial slice
          using a \texttt{jet} colourmap;
    \item a SHAP waterfall plot for the current patient alongside a beeswarm
          summary across the test cohort.
\end{itemize}

The application is packaged as a Docker~\cite{docker2020} container,
ensuring identical behaviour across development, evaluation, and deployment
environments. Per-prediction latency on commodity hardware (CPU-only
inference) is under two seconds with the TorchScript export and the
precomputed SHAP background.

%-----------------------------------------------------------------------
\section{Software Environment}
\label{sec:appB-software}

The implementation uses the following software stack. Exact versions are
pinned in the \texttt{requirements.txt} and \texttt{Dockerfile} that ship
with the project repository.

\begin{itemize}
    \item Python 3.10
    \item PyTorch 2.1 and torchvision~\cite{paszke2019pytorch}
    \item scikit-learn~\cite{pedregosa2011scikit}
    \item \texttt{pytorch-grad-cam}~\cite{selvaraju2017grad}
    \item \texttt{shap}~\cite{lundberg2017unified}
    \item SimpleITK~\cite{muschelli2014simpleitk}
    \item Streamlit 1.32~\cite{streamlit2021}
    \item Docker for containerisation~\cite{docker2020}
\end{itemize}

This appendix is intended to provide enough technical detail to reproduce
the preprocessing, model training, explainability, and dashboard deployment
described in the main thesis.