 \begin{center}\vspace{1.5in}
	\bfseries\LARGE{ACKNOWLEDGMENTS} \\
\end{center}\vspace{0.2in}
%\begin{acknowledgements}   
	   
	   
	  
\normalsize{\begin{spacing}{1.5}
% ------------------------------------------------------------------------
\noindent 
I would like to express my heartfelt gratitude to everyone who has contributed to make this thesis a memorable experience and has inspired this work in some way. Let me begin by expressing my gratitude to the Almighty God for the numerous blessings he has bestowed upon me.\\

We are happy to present this project after completing it successfully. This project would not
have been possible without the guidance, assistance and suggestions of many individuals. We
would like to express our deep sense of gratitude and indebtedness to each and everyone who
has helped us make this project a success.\\

We heartily thank Dr. Sanjay H A, Principal, BMS Institute of Technology \&
Management for his constant encouragement and inspiration in taking up this Project.\\

We heartily thank Dr. Anupama H S, HoD, Dept. of Artificial Intelligence and Machine Learning,
BMS Institute of Technology \& Management for her constant encouragement and
inspiration in taking up this project.\\

We gracefully thank our Project guide, Prof. Megha S, Asst. Professor,
Dept. of Artificial Intelligence and Machine Learning, for her encouragement and advice
throughout the course of the Project work.\\

Special thanks to all the staff members of Artificial Intelligence and Machine Learning Department for their help and
kind co-operation. \\

{\noindent
Bhoomika S Gowda\\
Devathi Saranya\\
Emma Kappen\\
Honey Hemant Hampannavar\\
\noindent
May 2026
}

%\vspace{1.5cm}
%Name of candidate
% ------------------------------------------------------------------------
\end{spacing}}
%\end{acknowledgements}
