\chapter{Introduction}\thispagestyle{empty}
\graphicspath{{Chapter1/figures/}}
{\setstretch{1.5}

\section{Context and Background}
Lung cancer is one of the most prevalent and fatal malignancies worldwide, accounting for a disproportionately high share of cancer-related deaths due to its characteristically late-stage presentation at the time of diagnosis. Early detection remains the most significant determinant of patient survival; however, identifying malignant pulmonary nodules in CT scans demands considerable radiological expertise and is prone to inter-observer variability. The rapid proliferation of artificial intelligence in medical imaging has opened new avenues for automating and augmenting diagnostic workflows, enabling faster, more consistent analysis of complex radiological data. Despite promising advances, most existing AI-based diagnostic systems function as opaque black-box models, producing predictions without interpretable reasoning — a critical barrier to clinical adoption. LungInsight was developed to bridge this gap by building a transparent, multimodal AI framework that combines CT imaging with structured clinical data to deliver both accurate and explainable lung cancer predictions.

\section{Purpose of the Project}
The primary objective of LungInsight is to develop a multimodal deep learning system capable of predicting the probability of lung cancer malignancy and classifying disease stage from CT scan images and clinical patient records. The project further aims to integrate Explainable AI techniques — specifically Grad-CAM for imaging and SHAP for clinical features — to ensure that model predictions are interpretable and clinically trustworthy. By delivering a Streamlit-based decision support dashboard, LungInsight seeks to provide clinicians with an accessible, real-time tool to support early diagnosis and evidence-based decision-making.

\section{Scope of the Project}
This project encompasses the design, development, and evaluation of a multimodal lung cancer prediction pipeline using the LIDC-IDRI CT dataset and the National Lung Screening Trial clinical records. The scope includes CT image preprocessing, clinical data preparation, model architecture design using DenseNet-based and MLP-based encoders with late fusion, explainability integration via Grad-CAM and SHAP, and comprehensive performance evaluation. The project does not extend to real-time integration with hospital information systems, regulatory approval for clinical deployment, or the diagnosis of lung conditions other than malignancy and stage classification.

\section{Significance of the Project}
Lung cancer's high mortality rate is closely linked to the absence of reliable, accessible, and explainable early diagnostic tools, particularly in settings with limited access to specialist radiologists. LungInsight addresses this gap by combining the predictive power of deep learning with the transparency of explainable AI, enabling clinicians to not only receive a diagnosis but also understand the imaging regions and clinical factors driving each prediction. The system has direct implications for SDG 3 (Good Health and Well-being) by improving survival outcomes through early detection, SDG 10 (Reduced Inequalities) by reducing dependence on specialist infrastructure, and SDG 9 (Industry, Innovation and Infrastructure) by advancing intelligent healthcare technology.

\section{System Architecture}
LungInsight adopts a multimodal late-fusion architecture comprising two parallel encoding branches. The imaging branch employs a DenseNet-based convolutional neural network pretrained for grayscale CT slice analysis to extract spatial features from pulmonary nodule regions. The clinical branch uses a fully connected multi-layer perceptron to encode tabular patient data — including age, sex, smoking history, emphysema score, and lung function — into a dense embedding. The outputs of both branches are concatenated and passed through fusion dense layers to produce a final malignancy probability and stage classification. Explainability is overlaid through Grad-CAM heatmaps on CT slices and SHAP value attribution on clinical features. The complete inference pipeline is exposed via a Streamlit dashboard for clinical use.

\section{Problem Statement}
Early-stage lung tumours are characteristically small and difficult to distinguish from benign nodules in CT scans, making manual diagnosis both time-intensive and susceptible to error. Existing AI diagnostic models, while achieving high classification accuracy, largely fail to provide interpretable reasoning for their predictions — making it difficult for clinicians to validate or trust AI-generated outputs. Furthermore, most systems rely solely on imaging data, disregarding the wealth of diagnostic signal present in structured clinical records. LungInsight addresses these limitations by proposing a unified, explainable, and multimodal prediction framework that is both clinically accurate and transparent.

\section{Objectives}
\begin{enumerate}
    \item Develop a multimodal lung cancer prediction model using a DenseNet-based CT image encoder and an MLP-based clinical feature encoder with late fusion, to jointly predict malignancy probability and disease stage.
    \item Integrate Explainable AI techniques — Grad-CAM for CT image localisation and SHAP for clinical feature attribution — to produce interpretable, patient-specific diagnostic explanations.
    \item Evaluate the framework comprehensively using accuracy, ROC-AUC, F1-score, PR-AUC, Brier score, calibration curves, and decision curve analysis, with ablation studies across imaging-only, clinical-only, and fused model configurations.
    \item Deploy a Streamlit-based clinical decision support dashboard enabling clinicians to upload CT scans and patient records, run inference, and review explainable risk assessments in real time.
\end{enumerate}

\section{Applications}
LungInsight has broad applicability across clinical and research settings. In hospital radiology departments, it can serve as a second-opinion tool, flagging high-risk CT scans for priority review. In under-resourced or rural healthcare settings where specialist radiologists are scarce, the system can extend diagnostic capability to general practitioners through its intuitive dashboard interface. The SHAP-based clinical explainability module is additionally valuable for oncology research, enabling identification of the most diagnostically significant patient-level risk factors. The framework's modular architecture also allows adaptation to other thoracic malignancies or imaging modalities with minimal modification.

\section{Structure of the Report}
This report is organised as follows: Chapter 2 presents an extended literature review of AI-based lung cancer detection methods, summarising key algorithms, datasets, results, and identified research gaps; Chapter 3 details the proposed methodology, including data acquisition, preprocessing pipelines, model architecture, training protocol, and explainability implementation; Chapter 4 describes the experimental setup and implementation environment; Chapter 5 presents the results and quantitative evaluation of the LungInsight framework; and Chapter 6 offers conclusions and directions for future work.

\begin{table}[H]
\centering
\captionsetup{justification=centering}
\caption{Report Organisation}
\begin{tabular}{clc}
\toprule
\textbf{Chapter No.} & \textbf{Content} & \textbf{Objective} \\
\midrule
1 & Introduction & — \\
\hline
2 & Literature Review of AI-Based Lung Cancer Detection & 1 \\
\hline
3 & Proposed Methodology: Multimodal Fusion \& XAI Pipeline & 1, 2 \\
\hline
4 & Implementation \& Experimental Setup & 3 \\
\hline
5 & Results \& Evaluation & 3 \\
\hline
6 & Conclusion \& Future Work & 4 \\
\bottomrule
\end{tabular}
\label{ReportOrg}
\end{table}

\textbf{Chapter 6 Conclusion} summarises the key findings of the LungInsight project, reflects on the effectiveness of the multimodal explainable AI approach for lung cancer diagnosis, and outlines directions for future research including multi-centre validation, integration with electronic health records, and extension to three-dimensional CT volume analysis.